Squintability and Other Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimization Choices

### 2025-11-18 revision/response.qmd

5. Some more practical aspects of how the new metrics are defined are given later in the paper (Fig. 5 and 6), in my opinion the illustrations in the Figures are not needed, but the practical details should be explained together with the definitions in section 3. For the smoothness metric the examples in Fig 2 and 3 are useful to get an intuition about the index, and something comparable would help with better understanding the squintability metric (probably some additional information on something like Fig 9 would help here).

\textcolor{blue}{We thank the reviewer for this insightful suggestion. In line with the reviewer’s recommendation, we have also strengthened Section 3.2 by adding a new figure (Figure 4) that provides intuition for the squintability metric. We have retained the two illustrative figures (now Figures 6 and 7) in Section 5.2, as we believe that some readers may still find the visualisation helpful when interpreting the practical behaviour of the metrics.}

6. The results in section 6 are very interesting. However, I was wondering about some details in 6.3: from the results shown it seems that the smoothness and squintability for the selected data examples are highly correlated, could this cause issues with the model fit shown in Table 2? It would be interesting if a similar result for CRS could be included, to see the effect of smoothness in that setting.

\textcolor{blue}{We have revised the text to include a comment on the correlation. It is moderate and does not practically affect the fit. We have also noted that it is not feasible to run CRS for the sine-wave optimization because it is not an efficient optimizer and fails to converge too often. On small dimensionality, it can be seen that it is not affected by lack of smoothness of an index. This would be expected because it is not a derivative-based optimizer, and, similarly to JSO small local fluctuations won't affect it's search.}

### 2025-11-18 jso.qmd
finished revising the paper and response about correlation and CRS for the sine-wave

@tbl-smoothness-squintability presents values computed for a subset of
PP indexes on the sine wave data for dimensions 4-8. (A full table of
the parameters estimated from the Gaussian process (outputscale $\eta$,
lengthscale $\ell$, smoothness $\nu$, and nugget $\sigma$) and from the
scaled logistic function ($\theta_1$ to $\theta_4$) for calculating
smoothness and squintability for all PP indexes and data combinations is
included in the supplementary materials). The column $\nu$ is used as
the smoothness metric and the column $\varsigma$ is calculated as
equation (@eq-squintability-parametric) as the squintability metric.

@@

In relation to the model fit, it should be noted that there is moderate correlation between squintability and smoothness, but this does not substantially affect the fit. Fitting smoothness alone results in this variable being weakly significant for explaining success rate, but the effect is dominated by squintability. The addition of interaction terms for squintability and data
dimension, and smoothness and data dimension, improves the model slightly, but not enough to be practically important or change the interpretation above.

This results support the use of the proposed
measures to effectively capture key factors affecting the success
rate of PP.

>>>>

@tbl-smoothness-squintability presents values computed for a subset of
PP indexes on the sine wave data for dimensions 4-8. (A full table of
the parameters estimated from the Gaussian process <!-- (outputscale $\eta$,
lengthscale $\ell$, smoothness $\nu$, and nugget $\sigma$)--> and from the
scaled logistic function <!--($\theta_1$ to $\theta_4$)--> for calculating
smoothness and squintability for all PP indexes and data combinations is
included in the supplementary materials). The column $\nu$ is used as
the smoothness metric and the column $\varsigma$ is calculated as
equation (@eq-squintability-parametric) as the squintability metric.

@@

In relation to the model fit, it should be noted that there is moderate correlation between squintability and smoothness, but this does not substantially affect the fit. Fitting smoothness alone results in this variable being weakly significant for explaining success rate, but the effect is dominated by squintability. The addition of interaction terms for squintability and data
dimension, and smoothness and data dimension, improves the model slightly, but not enough to be practically important or change the interpretation above. The CRS optimizer couldn't be used here because it is not efficient (@fig-proj) but we would expect that smoothness does not overly affect it either. Both JSO and CRS are expected to be robust to small local fluctuations.

These results also support the use of the proposed
measures to effectively capture key factors affecting the success
rate of PP.

### 2025-11-18 jso.tex
finished revising the paper and response about correlation and CRS for the sine-wave

\begin{tcolorbox}[enhanced jigsaw, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, breakable, coltitle=black, title={Algorithm: Jellyfish Optimizer Pseudo Code}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, left=2mm, colback=white, bottomtitle=1mm, opacityback=0, toprule=.15mm, toptitle=1mm, titlerule=0mm, rightrule=.15mm, leftrule=.75mm, opacitybacktitle=0.6]

@@

Table~\ref{tbl-smoothness-squintability} presents values computed for a
subset of PP indexes on the sine wave data for dimensions 4-8. (A full
table of the parameters estimated from the Gaussian process (outputscale
\(\eta\), lengthscale \(\ell\), smoothness \(\nu\), and nugget
\(\sigma\)) and from the scaled logistic function (\(\theta_1\) to
\(\theta_4\)) for calculating smoothness and squintability for all PP
indexes and data combinations is included in the supplementary
materials). The column \(\nu\) is used as the smoothness metric and the
column \(\varsigma\) is calculated as equation
(\ref{eq-squintability-parametric}) as the squintability metric.

@@

In relation to the model fit, it should be noted that there is moderate
correlation between squintability and smoothness, but this does not
substantially affect the fit. Fitting smoothness alone results in this
variable being weakly significant for explaining success rate, but the
effect is dominated by squintability. The addition of interaction terms
for squintability and data dimension, and smoothness and data dimension,
improves the model slightly, but not enough to be practically important
or change the interpretation above.

This results support the use of the proposed measures to effectively
capture key factors affecting the success rate of PP.

>>>>

\begin{tcolorbox}[enhanced jigsaw, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, title={Algorithm: Jellyfish Optimizer Pseudo Code}, breakable, bottomtitle=1mm, leftrule=.75mm, titlerule=0mm, toprule=.15mm, left=2mm, colback=white, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, arc=.35mm, coltitle=black, toptitle=1mm, rightrule=.15mm, opacitybacktitle=0.6]

@@

Table~\ref{tbl-smoothness-squintability} presents values computed for a
subset of PP indexes on the sine wave data for dimensions 4-8. (A full
table of the parameters estimated from the Gaussian process and from the
scaled logistic function for calculating smoothness and squintability
for all PP indexes and data combinations is included in the
supplementary materials). The column \(\nu\) is used as the smoothness
metric and the column \(\varsigma\) is calculated as equation
(\ref{eq-squintability-parametric}) as the squintability metric.

@@

In relation to the model fit, it should be noted that there is moderate
correlation between squintability and smoothness, but this does not
substantially affect the fit. Fitting smoothness alone results in this
variable being weakly significant for explaining success rate, but the
effect is dominated by squintability. The addition of interaction terms
for squintability and data dimension, and smoothness and data dimension,
improves the model slightly, but not enough to be practically important
or change the interpretation above. The CRS optimizer couldn't be used
here because it is not efficient (Figure~\ref{fig-proj}) but we would
expect that smoothness does not overly affect it either. Both JSO and
CRS are expected to be robust to small local fluctuations.

These results also support the use of the proposed measures to
effectively capture key factors affecting the success rate of PP.
